Review: SASB SAT Solving for Vesicle Traffic Systems in Cells

### 2017-08-16 paper/encoding.tex
changes
sasb-update

We will describe the Boolean constraints that encode VTSs in several categories.
%
In the end, we will present in a table the constraints needed for the
model variants.
%
To avoid cumbersome notation, we will not explicitly write the ranges of the indexing
in the constraints.
%
$i$ and $j$ will range over nodes.
%
$m$ will range over molecules.
%
$q$ will range over edges between two nodes.
%

% \subsubsection{Constraints on the structure of the edges, nodes and }
\subsubsection{Constraints on presence, activity of the molecule, and pairing matrix}
%
We need the following constraints~\eqref{eq:f0} and~\eqref{eq:c3}
to encode the basic structure of VTSs.
%
For an edge to exist it should have one molecule present. 
%
\begin{align}
  \bigwedge\limits_{i,j,q} (\bigvee_m e_{i,j,q,m} )\limplies e_{i,j,q}\tag{V1}\label{eq:f0}
\end{align}
If a molecule is active on an edge, it should be present on the edge.
%
\begin{align}
  \bigwedge\limits_{i,j,q,m} b_{i,j,q,m} \limplies e_{i,j,q,m}\tag{V2}\label{eq:f1}
\end{align}
A molecule should be present to be active on a node.  
\begin{align}
  \bigwedge\limits_{i,m} a_{i,m} \limplies n_{i,m}
  \tag{V3}\label{eq:c4}
\end{align}
The edge labels are subset of the node labels of the source and target nodes.
\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies (n_{i,m} \land n_{j,m} )\tag{V4}\label{eq:c0}
\end{align}
Self edges are not allowed. 
\begin{align}
   \bigwedge\limits_{i,q} \neg e_{i,i,q}\tag{V5}\label{eq:c2}
\end{align}
We fix first half as Q-SNAREs and rest as R-SNAREs and set diagonal blocks in pairing matrix to be 0's.
%It reduces the search space.
% Condition on $p_{kk'}$. \ashu{why???} .
\begin{align}
  \bigwedge\limits_{(x < M/2 \, \land  \, y < M/2) \lor  (x >= M/2 \land y >= M/2)} \neg p(x,y)
  \tag{V6}\label{eq:c3}
\end{align}

@@

\subsubsection{Regulation on nodes and edges}
We are considering several models that differ in constraints on
the activity of molecules.
%
We will present~\eqref{eq:ann}-~\eqref{eq:aep} that encodes
the varying constraints.
%
The following constraint encodes no conditions on activities on nodes,
i.e., all the present molecules on the nodes are active.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} = a_{i,m}    \tag{Ann}\label{eq:ann}
\end{align}
The following constraint encodes that activity of a molecule $m$ on the node is
defined by a Boolean function $\nodef_m$ of presence of molecules present on that node.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} \limplies a_{i,m} =  \nodef_m (n_{i,1},\dots,n_{i,\mu}) 
\tag{Anb}\label{eq:anb}
\end{align}
The following constraint encodes no conditions on activities on edges,
i.e., all the present molecules on the edges are active.
\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} = b_{i,j,q,m}
\tag{Aen}\label{eq:aen}
\end{align}
The following constraint encodes that activity of a molecule $m$ on the edge is
defined by a Boolean function $\edgef_m$ of presence of molecules present on that edge.
\begin{align}
   \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies b_{i,j,q,m} = \edgef_k(e_{i,j,q,1}, .., e_{i,j,q,\mu} )
  \tag{Aeb}\label{eq:aeb}
\end{align}
%
The following constraint encodes that the activity of the molecules on
edges is defined by inhibition by other molecules based on the pairing
matrix. 
\begin{align}
   \bigwedge\limits_{i,j,q, m}  [e_{i,j,q,m} \limplies  \bigwedge_{m' \neq m} (p_{m,m'} \limplies e_{i,j,q,m'})] \lequiv (\neg b_{i,j,q,m} \land\;\;\;\;  \bigwedge_{\mathclap{m' \neq m, p_{m,m'}}} \neg b_{i,j,q,m'})
  \tag{Aep}\label{eq:aep}
\end{align}
%
\subsubsection{Constraints for stability condition}
%
We use $m$-reachability to encode the stability condition in VTSs.
%
The following constraint recursively encodes that node $j$ is $m$-reachable from node $i$ in less than $p$ steps
if either there is a direct edge between $i$ and $j$ with $m$ present on the edge or there is a edge between $i^{\prime \prime}$ and
$(i \neq i^{\prime \prime})$ with $m$ present, and j is $m$-reachable from $i'$ in less than $p-1$ steps.
%
\begin{align}
  \bigwedge\limits_{i,j,m,p} r_{i,j,m,p} \limplies (\bigvee_{q} \, e_{i,j,q,m} \lor \bigvee_{i\neq i^{\prime}} ( \, \bigvee_{q} e_{i,i^{\prime},q,m}) \land r_{i^{\prime},j,m,p-1} )
  \tag{R1}\label{eq:reach1}
\end{align}
Now we can encode stability using the reachability variables.
and say if there is an $m$-edge between $i$ and $j$, there is
$m$-reachable path from $j$ to $i$.
\begin{align}
 \bigwedge\limits_{i,j,m} (\bigvee_{q} e_{i,j,q,m}) \limplies r_{j,i,m,|N|}
  \tag{R2}\label{eq:reach2}
\end{align}

@@

%if all the pairs of the present molecule are present on the same edge than all of them inhibit each other resulting in making all the molecules inactive, we term this process ``pairing inhibition".
%
The constraints for the variants can be given to a SAT solver to find
VTSs that belong to the variants.

>>>>

We will describe the Boolean constraints that encode VTSs in several categories.
%
In the end, we will present in a table the constraints needed for the
model variants.
%
To avoid cumbersome notation, we will not explicitly write the ranges of the indexing
in the constraints.
%
$i$ and $j$ will range over nodes.
%
$m$ will range over molecules.
%
$q$ will range over edges between two nodes, i.e., from $1$ to $\pi$.
%

\subsubsection{Constraints on presence, activity of the molecule, and pairing matrix}
%
We need the following constraints~\eqref{eq:f0} and~\eqref{eq:c3}
to encode the basic structure of VTSs.
%
For an edge to exist it should have one molecule present. 
%
\begin{align}
  \bigwedge\limits_{i,j,q} (\bigvee_m e_{i,j,q,m} )\limplies e_{i,j,q}\tag{V1}\label{eq:f0}
\end{align}
If a molecule is active on an edge, it should be present on the edge.
%
\begin{align}
  \bigwedge\limits_{i,j,q,m} b_{i,j,q,m} \limplies e_{i,j,q,m}\tag{V2}\label{eq:f1}
\end{align}
A molecule should be present to be active on a node.  
\begin{align}
  \bigwedge\limits_{i,m} a_{i,m} \limplies n_{i,m}
  \tag{V3}\label{eq:c4}
\end{align}
The edge labels are subset of the node labels of the source and target nodes.
\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies (n_{i,m} \land n_{j,m} )\tag{V4}\label{eq:c0}
\end{align}
Self edges are not allowed. 
\begin{align}
   \bigwedge\limits_{i,q} \neg e_{i,i,q}\tag{V5}\label{eq:c2}
\end{align}
We fix first half as Q-SNAREs and rest as R-SNAREs and set diagonal blocks in pairing matrix to be 0's.
%It reduces the search space.
% Condition on $p_{kk'}$. \ashu{why???} .
\begin{align}
  \bigwedge\limits_{(x < M/2 \, \land  \, y < M/2) \lor  (x >= M/2 \land y >= M/2)} \neg p(x,y)
  \tag{V6}\label{eq:c3}
\end{align}

@@

\subsubsection{Regulation on nodes and edges}
We are considering several models that differ in constraints on
the activity of molecules.
%
We will present~\eqref{eq:ann}-~\eqref{eq:aep} that encodes
the varying constraints.
%
The following constraint encodes no conditions on activities on nodes,
i.e., all the present molecules on the nodes are active.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} = a_{i,m}    \tag{Ann}\label{eq:ann}
\end{align}
The following constraint encodes that activity of a molecule $m$ on the node is
defined by a Boolean function $\nodef_m$ of presence of molecules present on that node.
\begin{align}
\bigwedge\limits_{i,m} n_{i,m} \limplies a_{i,m} =  \nodef_m (n_{i,1},\dots,n_{i,\mu}) 
\tag{Anb}\label{eq:anb}
\end{align}
The following constraint encodes no conditions on activities on edges,
i.e., all the present molecules on the edges are active.
\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} = b_{i,j,q,m}
\tag{Aen}\label{eq:aen}
\end{align}
The following constraint encodes that activity of a molecule $m$ on the edge is
defined by a Boolean function $\edgef_m$ of presence of molecules present on that edge.
\begin{align}
   \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies b_{i,j,q,m} = \edgef_k(e_{i,j,q,1}, .., e_{i,j,q,\mu} )
  \tag{Aeb}\label{eq:aeb}
\end{align}
%
The following constraint encodes that the activity of the molecules on
edges is defined by inhibition by other molecules based on the pairing
matrix. 
\begin{align}
   \bigwedge\limits_{i,j,q, m}  [e_{i,j,q,m} \limplies  \bigwedge_{m' \neq m} (p_{m,m'} \limplies e_{i,j,q,m'})] \lequiv (\neg b_{i,j,q,m} \land\;\;\;\;  \bigwedge_{\mathclap{m' \neq m, p_{m,m'}}} \neg b_{i,j,q,m'})
  \tag{Aep}\label{eq:aep}
\end{align}
%
\subsubsection{Constraints for stability condition}
%
We use $m$-reachability to encode the stability condition in VTSs.
%
The following constraint recursively encodes that node $j$ is $m$-reachable from node $i$ in less than $p$ steps
if either there is a direct edge between $i$ and $j$ with $m$ present on the edge or there is a edge between $i^{\prime \prime}$ and
$(i \neq i^{\prime \prime})$ with $m$ present, and j is $m$-reachable from $i'$ in less than $p-1$ steps.
%
\begin{align}
  \bigwedge\limits_{i,j,m,p} r_{i,j,m,p} \limplies (\bigvee_{q} \, e_{i,j,q,m} \lor \bigvee_{i\neq i^{\prime}} ( \, \bigvee_{q} e_{i,i^{\prime},q,m}) \land r_{i^{\prime},j,m,p-1} )
  \tag{R1}\label{eq:reach1}
\end{align}
Now we can encode stability using the reachability variables.
and say if there is an $m$-edge between $i$ and $j$, there is
$m$-reachable path from $j$ to $i$.
\begin{align}
 \bigwedge\limits_{i,j,m} (\bigvee_{q} e_{i,j,q,m}) \limplies r_{j,i,m,\nu}
  \tag{R2}\label{eq:reach2}
\end{align}

@@

%if all the pairs of the present molecule are present on the same edge than all of them inhibit each other resulting in making all the molecules inactive, we term this process ``pairing inhibition".
%
The constraints for the variants can be given to a SMT solver to find
VTSs that belong to the variants.

### 2017-08-16 review-sasb-sat/main.tex
sasb-update

\section{Review 1}
\begin{enumerate}
\item {Overall evaluation: Weak accept\\}
\end{enumerate}
\textbf{{\color{black} Evaluation\\}}
{\color{green!20!blue}The paper describes a SAT encoding of the model identification problem for graphical
models of cellular vesicle traffic systems (VTS).
By integrating all the constraints on admissible graphs for VTS, and in particular
regarding stability and reachability constraints, the author obtained an efficient
implementation relying on Z3 which outputs a model (if exists) which satisfy
necessary constraints for VTS and variants of additional constraints on the
topology/functionality of the model.\\}

As far as I understood, the method allows to identify one graph (if any) ; \todo{Discuss with Ashutosh} \textbf{it would be interesting to consider counting the number of solutions as well.}

The paper is very technical and focuses mainly on the SAT encoding of the problem
and \textbf{gives very few examples of application on concrete biological examples.} \todo{Mukund should address this} It will be interesting to discuss the results of the different variants of the model identification on small cases with a biological perspective.\\

@@

{\color{green!20!blue}The paper presents a SAT-based method to study vesicle traffic systems and
specifically, to derive network models for these systems that satisfy some given
properties, specifically, k-connectedness.

The motivation behind the work is very important, i.e. understanding the regulation
of the SNARE protein that is crucial in this system and in particular, in
determining specific network topologies. \todo {We should motivate why k-connected is important? Just reference to old paper i think is not enough. I and Mukund can address this.} \textbf{However biological considerations are only
marginal in this work, and the use of the k-connectedness property} (the only
property considered in this work) \textbf{is not motivated or justified}. \textbf{The authors should
put more effort into explaining why k-connectedness is important here. }}

Discussion of results also reflect this \todo {we should discuss this} \textbf{lack of bio insights}, and is mainly devoted to comparing performance w.r.t. a previous BMC-based version of the approach.

On the other hand, the SAT-based method is adequate, sound and very relevant (for
both this case study and SASB), even though the encoding is quite straightforward.
The paper is generally well written and easy to follow. \\

Please find more detailed comments below:\\
\begin{enumerate}
\item one sentence of the abstract is misleading: "what are all possible networks for
various combinations of those properties?" what you actually do is finding ONE
network (i.e. a model satisfying the SMT problem), not all networks. Consider
rephrasing this.\newline
\todo{We should discuss how to rephrase this or stay as it is.}
{\color{red}{Mukund, Ashutosh}}\\
\item can you better motivate the steady state assumption? have you performed any
dynamic/transient analysis of such derived networks? \newline
{\color{red}{Mukund}}\\
\item Page 3, Line 110: "in which a two SNAREs" --"in which two SNAREs" \newline{\color{red}{Fixed!}}\\

\item  Page 3, "synthesis" paragraph: \todo{Not sure why people are finding too many biology section gaps, happened with last paper also. May be they are too much interested in understanding the biology.} I don't get what you are explaining here, is that a list of analysis performed? I couldn't see any of this in the result section. \newline
{\color{red}{Mukund}}\\

>>>>

\section{Review 1}
\begin{enumerate}
\item {Overall evaluation: Weak accept\\}
\end{enumerate}
\textbf{{\color{black} Evaluation\\}}
The paper describes a SAT encoding of the model identification problem for graphical
models of cellular vesicle traffic systems (VTS).
By integrating all the constraints on admissible graphs for VTS, and in particular
regarding stability and reachability constraints, the author obtained an efficient
implementation relying on Z3 which outputs a model (if exists) which satisfy
necessary constraints for VTS and variants of additional constraints on the
topology/functionality of the model.\\

As far as I understood, the method allows to identify one graph (if any) ; \textbf{it would be interesting to consider counting the number of solutions as well.}\newline
{\color{red}{Ashutosh}}

The paper is very technical and focuses mainly on the SAT encoding of the problem
and \textbf{gives very few examples of application on concrete biological examples.}\\ 
{\color{red}{Ankit\\}}

It will be interesting to discuss the results of the different variants of the model identification on small cases with a biological perspective.\\
{\color{red}{Our Comment/Reply:}}
%The reviewer has highlighted that it would be interesting to connect thedifferent variants of the model to relevant biological cases.
{\color{blue}{ We agree wholeheartedly. However, there is at the moment very little known about how
the regulation of SNAREs functions. What we have done here is to study many
possible variants of SNARE regulation, from simple to complex. It is
difficult to say which one of these variants is likely to hold in real
cells, and indeed, different species might use different versions of SNARE
regulation. Luckily, even given the most general variant of SNARE
regulation (of which all other models are just special cases) we have still
managed to find a constraint on graph connectedness. Thus we are confident
that our results apply to existing cell-biological data. One of the authors
(MT) is presently assembling a list of known SNARE fluxes in yeast and
mammalian cells to see whether our predictions hold. This will be published
as a follow-up manuscript, and we are happy to share a draft of this
manuscript with the reviewers.}}

@@

The paper presents a SAT-based method to study vesicle traffic systems and
specifically, to derive network models for these systems that satisfy some given
properties, specifically, k-connectedness.

The motivation behind the work is very important, i.e. understanding the regulation
of the SNARE protein that is crucial in this system and in particular, in
determining specific network topologies. \textbf{However biological considerations are only
marginal in this work, and the use of the k-connectedness property} (the only
property considered in this work) \textbf{is not motivated or justified}. \textbf{The authors should
put more effort into explaining why k-connectedness is important here. }

{\color{red}{Fixed! refer new updated draft.\\}}

Discussion of results also reflect this \textbf{lack of bio insights}, and is mainly devoted to comparing performance w.r.t. a previous BMC-based version of the approach. \newline 
{\color{red}{Commented! see first review blue color paragraph\\}}

On the other hand, the SAT-based method is adequate, sound and very relevant (for
both this case study and SASB), even though the encoding is quite straightforward.
The paper is generally well written and easy to follow. \\ \\

Please find more detailed comments below:\\
\begin{enumerate}
\item one sentence of the abstract is misleading: "what are all possible networks for
various combinations of those properties?" what you actually do is finding ONE
network (i.e. a model satisfying the SMT problem), not all networks. Consider
rephrasing this.\newline
\todo{We should discuss how to rephrase this or stay as it is.}
{\color{red}{Ashutosh.}}\\

\item can you better motivate the steady state assumption? have you performed any
dynamic/transient analysis of such derived networks? \newline
{\color{red}{Fixed ! refer updated draft.}}\\

\item Page 3, Line 110: "in which a two SNAREs" --"in which two SNAREs" \newline{\color{red}{Fixed!}}\\

\item  Page 3, "synthesis" paragraph: I don't get what you are explaining here, is that a list of analysis performed? I couldn't see any of this in the result section. \newline
{\color{red}{Fixed! See updated draft}}\\
